---
title: "DOH Exam"
author: "Me"
date: "3/31/2021"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, error = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(lubridate)
library(arsenal)
library(data.table)
library(tidyr)
library(broom)
library(knitr)
```


# Explore the data


Load datasets
```{r}
june5 = read_csv('Exam List June 5.csv', na = c('','-','NA'))
june6 = read_csv('Exam List June 6.csv', na = c('','-','NA'))
june7 = read_csv('Exam List June 7.csv', na = c('','-','NA'))
```

Check column names
```{r}
colnames(june5)
```

June 5 has no columns drucity and druprovince
```{r}
colnames(june6)[!(colnames(june6) %in% colnames(june5))]
```


Add these columns and set to NA
```{r}
june5$drucity = NA
june5$druprovince = NA
```



```{r}
glimpse(june7)
```

Im assuming 1 case number per patient. Check if this is true
```{r}
june5 %>%
        group_by(casenumber) %>%
        count() %>%
        filter(n > 1, !is.na(casenumber))
```

```{r}
june6 %>%
        group_by(casenumber) %>%
        count() %>%
        filter(n > 1, !is.na(casenumber))
```
Same multiple entries on June 7


Check this casenumbers
```{r}
multiple_cases = june5 %>%
        group_by(casenumber) %>%
        count() %>%
        filter(n > 1, !is.na(casenumber)) %>%
        pull(casenumber)

june5 %>%
        filter(casenumber %in% multiple_cases) %>%
        arrange(casenumber) %>%
        head(15) %>%
        kable()
```
Focusing only on the demographics

- PH10519 is both female and male,
- PH11142 changed age
- PH12184 is both chinese and filipino
- PH2315 is both age 53 and 23


Try to check these casenumbers in june 7 data
```{r}
june7 %>%
        filter(casenumber %in% multiple_cases) %>%
        arrange(casenumber) %>%
        distinct(casenumber, sex, age, nationality, .keep_all = F) %>%
        group_by(casenumber) %>%
        add_count() %>%
        filter(n > 1) %>%
        head(15) %>%
        kable()
```

It seems that this was not updated in the june 7 data and looking at the demographics alone, there are a lot of inconsistencies.



Check the completeness of dates columns and the dates columns
```{r}

june5 %>%
        select(ends_with('_date')) %>%
        map_dbl(function(x) sum(is.na(x)/ length(x)))
```

Looks like the specimen date has the least missing data

```{r}
june5 %>%
        select(contains('date')) %>%
        arrange(mdy(specimen_date)) %>%
        head(15) %>%
        kable()
```

There are specimen dates that are way before 2020. There are also a lot of 1/1/2020 specimen date which I assume are entry errors. Im trying to look for ways to complete missing case numbers.


Check for june5 cases that were removed
```{r}
june5 %>%
        filter(!(casenumber %in% june6$casenumber)) %>%
        kable()
```

```{r}
june5 %>%
        filter(!(casenumber %in% june7$casenumber)) %>%
        kable()
```

There were 3 cases that were in june 5 but not in june 6 and 4 cases that are in june 5 but not in june 7. If they are just updating. This shouldnt be removed.


Check if there are some changes in some column values from June 5 to June 7.

Im interested whether there are cases that changed province and hospitals. This information is very important on whether to impose GCQ or ECQ on certain locations

```{r}
june7 %>%
        select(casenumber, currentregion) %>%
        left_join(june5 %>% select(casenumber, currentregion), by = 'casenumber') %>%
        distinct() %>%
        group_by(casenumber) %>%
        filter(currentregion.x != currentregion.y, !is.na(casenumber))
```

There are few cases where the patient changed current region


I am also interested whether there are some patients that were declared dead on june 5 but then has different condition on june 6 or 7
```{r}
june5 %>%
        select(casenumber, healthstatus) %>%
        filter(healthstatus == 'Died') %>%
        left_join(june7 %>% select(casenumber, healthstatus), by = c('casenumber' = 'casenumber')) %>%
        filter(!is.na(casenumber)) %>%
        filter(healthstatus.x != healthstatus.y)
```

At least there are no dead patients in June 5 that were alive in June 7 data

How about if some patients changed their gender
```{r}
june7 %>%
        select(casenumber, sex) %>%
        left_join(june5 %>% select(casenumber, sex), by = 'casenumber') %>%
        distinct() %>%
        group_by(casenumber) %>%
        filter(sex.x != sex.y, !is.na(casenumber)) %>%
        rename(sexjune5 = sex.y, sexjune7 = sex.x)
```

Sanity check
```{r}
changed_gender = june7 %>%
        select(casenumber, sex) %>%
        left_join(june5 %>% select(casenumber, sex), by = 'casenumber') %>%
        distinct() %>%
        group_by(casenumber) %>%
        filter(sex.x != sex.y, !is.na(casenumber)) %>%
        rename(sexjune5 = sex.y, sexjune7 = sex.x) %>%
        pull(casenumber)

june5 %>%
        filter(casenumber %in% changed_gender) %>%
        arrange(casenumber) %>%
        select(casenumber, age, sex, nationality)
```
```{r}
june7 %>%
        filter(casenumber %in% changed_gender) %>%
        arrange(casenumber) %>%
        select(casenumber, age, sex, nationality)
```
Either multiple entry or just changed gender.

Compare if some missing values are filled on June 7 data. My assumption is June 7 has the most complete data
```{r}
june6_missing = aggregate(. ~ casenumber, data = june6, function(x) sum(is.na(x)), na.action = na.pass)

june6_missing = june6_missing %>%
        gather(-casenumber, key = 'col', value = 'count_missing') %>%
        group_by(casenumber) %>%
        summarise(missing_values = sum(count_missing))
```



```{r}
june7_missing = aggregate(. ~ casenumber, data = june7, function(x) sum(is.na(x)), na.action = na.pass)


june7_missing = june7_missing %>%
        gather(-casenumber, key = 'col', value = 'count_missing') %>%
        group_by(casenumber) %>%
        summarise(missing_values = sum(count_missing))
```

Rename columns
```{r}
june6_missing = june6_missing %>%
        rename_with(function(x) paste0('June6_', x))

june7_missing = june7_missing %>%
        rename_with(function(x) paste0('June7_', x))
```


```{r}
june7_missing %>%
        inner_join(june6_missing, by = c('June7_casenumber' = 'June6_casenumber')) %>%
        filter(June7_missing_values != June6_missing_values)
```

It seems that some missing value in june6 were now filled. Can assume that June7 is more updated

Check if there are entries in June7 that has more missing values
```{r}
more_missing = june7_missing %>%
        inner_join(june6_missing, by = c('June7_casenumber' = 'June6_casenumber')) %>%
        filter(June7_missing_values > June6_missing_values)

june7 %>%
        filter(casenumber %in% more_missing$June7_casenumber) %>%
        select(casenumber,datespecimenreceivedbylaboratory)
```

```{r}
june6 %>%
        filter(casenumber %in% more_missing$June7_casenumber) %>%
        select(casenumber,datespecimenreceivedbylaboratory)
```

Okay, some entries for the datespecimenreceivedbylaboratory were changed from unknown date format to NA


Check entries in June 7 that were not in june5
```{r}
june7 %>%
        anti_join(june5, by = c('casenumber' = 'casenumber')) %>%
        summarise(earliest_date = min(mdy(result_date), na.rm = T), 
                  latest_date = max(mdy(result_date), na.rm = T))
```


```{r}
june5 %>%
        filter(year(mdy(result_date)) == '2020') %>%
        summarise(latest_date = max((mdy(result_date)), na.rm = T))
```

Okay that's weird. There's a December 2020 case in June5 data

```{r}
june5 %>%
        filter(mdy(result_date) == '2020-12-05') %>%
        kable()
```

This is probably May 12 2020

There are also inconsistencies in the date format. Some are Month day year and some are day month year. These inconsistencies is problematic for time-series and time related analysis

I am comparing the dates of June 5 and June 7 data to check whether the June7 data were new entries. Maybe just check the case number

New entries for June 7
```{r}
june7 %>%
        anti_join(june5, by = c('casenumber' = 'casenumber')) %>%
        arrange(desc(parse_number(casenumber))) %>%
        select(casenumber, age, sex, nationality, healthstatus)
```
There are 443 case numbers that are new on June 7

```{r}
june5 %>%
        arrange(desc(parse_number(casenumber))) %>%
        select(casenumber, age, sex, nationality, healthstatus) %>%
        head(15)
```

Im not sure how these new entries were gathered since the case number and dates or overlapping.


I saw one entry without a number in case number

```{r}
june7 %>%
        filter(!is.na(casenumber)) %>%
        filter(!(str_detect(casenumber,'[:digit:]'))) %>%
        kable()
```

# Combine the data

There are a lot of inconsistencies in the data. Maybe just assume that the June 7 is the most updated data. The goal now is to update entries from June 5 and June 6 using June 7 data. Also, make sure that the missing entries in June 7 that are in June 5 and 6 are also included. Some missing values in June 6 were  updated in the June 7. ASide from missing values, some values also changed in the June 7 entry. We will prioritize the values of June7. I think it is also safer not to remove multiple entries as one of those entries is possibly the right one.  Again, the assumption here is that the June 7 data is the most accurate data. Then clean the data after.


```{r}
june7 %>%
        filter(is.na(casenumber)) %>%
        nrow(.)
```
~5.3k rows without data


```{r}
#common values to all
common_to_all = june7 %>%
        semi_join(june6) %>%
        semi_join(june5)

#in June 7 but not in june 5 and 6
new_values_june7 = june7 %>%
        anti_join(june6) %>%
        anti_join(june5)


#with case number not in june 6 and 7
cases_in_june5_notin_june67 = june5 %>%
        anti_join(june6, by = c('casenumber')) %>%
        anti_join(june7, by = c('casenumber'))

#with case number in june6 not in june 7
cases_injune6_notin_june7 = june6 %>%
        anti_join(june7, by = c('casenumber'))

# bind all dataframe
june7_initial = bind_rows(common_to_all, new_values_june7, cases_in_june5_notin_june67, cases_injune6_notin_june7)

# get the remaining entries from june7
june7_combined = june7 %>%
        anti_join(june7_initial) %>%
        rbind(june7_initial)
```

The challenge now is how to get those without case number in June 5 and June6 but not recorded in June 7



```{r}
june6 %>%
        anti_join(june7_combined) %>%
         map_dbl(function(x) sum(is.na(x)/ length(x))) %>%
        as.data.frame() %>%
        rename(percent_missing = '.') %>%
        arrange(percent_missing)
```





```{r}
june5 %>%
        anti_join(june7_combined, by = c('casenumber')) %>%
        nrow(.)
```
```{r}
june6 %>%
        anti_join(june7_combined, by = c('casenumber')) %>%
        nrow(.)
```

Looking at case number alone, all data from June 5 and June6 were already at the june7 combined data

```{r}
june6 %>%
        anti_join(june7_combined) %>%
        filter(is.na(casenumber)) %>%
        head(10) %>%
        kable()
```
However there are cases like this where there is no case number to anchor the search. It may be possible that these patients were already recorded and updated in the June7 Combined data


Just assume a new patient based on this columns if casenumber is missing
- age
- sex
- nationality
- confirming_lab
- result_date

June5 assumed patients not in june7 combined
```{r}
#30 patients
assumed_june5_patients = june5 %>%
        filter(is.na(casenumber)) %>%
        anti_join(june7_combined, by = c('age','sex','nationality','confirminglab','result_date'))

#18 patients
assumed_june6_patients = june6 %>%
        filter(is.na(casenumber)) %>%
        anti_join(june7_combined, by = c('age','sex','nationality','confirminglab','result_date'))
```

The code above will identify possible patients that were not included in the june7 combined base on the age, sex, nationality, confirming lab, and result data

```{r}
all_entries = bind_rows(june7_combined, assumed_june5_patients, assumed_june6_patients)
all_entries = all_entries %>%
        distinct()
```


Check

```{r}
all_entries %>%
        anti_join(june7_combined) %>%
         map_dbl(function(x) sum(is.na(x)/ length(x))) %>%
        as.data.frame() %>%
        rename(percent_missing = '.') %>%
        arrange(percent_missing)
```
Almost same percentage of missing data


# Clean some data

Try to clean some data.


Nationality

```{r}
all_entries %>%
        count(nationality)
```

Not sure if current location is a good indicator of nationality. Might be a foreigner staying in PH
```{r}
all_entries = all_entries %>%
        mutate(nationality = str_to_title(nationality)) %>%
        mutate(nationality = str_replace(nationality, '^Australia$','Australian'),
               nationality = str_replace(nationality, '^India$','Indian'),
               nationality = str_replace(nationality, 'United Kingdom','British, Uk'),
               nationality = str_replace(nationality, '^Filipino$','Philippine, Filipino'))

#if case number starts with PH
all_entries = all_entries %>%
        mutate(nationality = ifelse(str_detect(casenumber, 'PH'),'Philippine, Filipino', nationality )) 

```


No Typo on health status
```{r}
all_entries %>%
        count(healthstatus)
```

```{r}
all_entries %>%
        select(onset_date) %>%
        mutate(onset_date = mdy(onset_date)) %>%
        arrange(onset_date)
```

Im not sure if this is default or not.

```{r}
all_entries %>%
        select(casenumber, age,specimen_date, onset_date, result_date) %>%
        mutate(specimen_date = mdy(specimen_date)) %>%
        arrange(specimen_date)
```
Some input their birthday on specimen date. I think result date is much accurate.

Try to fix specimen date. If the year is below 2020 just change it to result date.

```{r}
all_entries = all_entries %>%
        mutate(specimen_date = ifelse(year(mdy(specimen_date)) < '2020', result_date, specimen_date))
```


```{r}
all_entries %>%
        select(casenumber, result_date, datespecimenreceivedbylaboratory) %>%
        mutate(result_date = mdy(result_date)) %>%
        arrange((result_date))
```
There are few typos in result date. Try to fix

- Remove the 1970 entry
- Change the last digit to 2020
- If the date is greater than August, use day-month-year format instead
```{r}
all_entries = all_entries %>%
        mutate(result_date = ifelse(mdy(result_date)< '2000-01-01',NA, result_date)) %>%
        mutate(result_date = str_replace(result_date, '/[:digit:]{4}', '/2020')) %>%
        mutate(result_date = if_else(mdy(result_date) < '2020-08-01', mdy(result_date), dmy(result_date)))
       
#return to original format
all_entries = all_entries %>%
        mutate(result_date = format(result_date, format = '%m/%d/%Y'))
```


There are other date columns and I would probably do the same checking with the other date columns. The data wrangling will depend on how the data is collected, what system are they using, are there default values, etc


Check columns for symptoms
```{r}
all_entries %>%
        select(comorbidity:none, -specifycomorbidity) %>%
        stack() %>%
        rename(answer= values, column = ind) %>%
        count(answer)
```

Just some typo

Fix typo for those boolean answers
```{r}
all_entries = all_entries %>%
        mutate_at(c('comorbidity','fever','cough','cold','sorethroat','difficultyofbreathing','diarrhea','none','quarantined'), str_to_title) 
```



###############################
Create a function to combine and clean data

```{r}
library(tidyverse)
library(lubridate)
library(arsenal)
library(data.table)
library(tidyr)
library(broom)

##load the data

june5 = read_csv('Exam List June 5.csv', na = c('','-','NA'))
june6 = read_csv('Exam List June 6.csv', na = c('','-','NA'))
june7 = read_csv('Exam List June 7.csv', na = c('','-','NA'))

combine_data = function(data1, data2, data3){
        data1$drucity = NA
        data1$druprovince = NA
        
        
                #common values to all
        common_to_all = data3 %>%
                semi_join(data2) %>%
                semi_join(data1)
        
        #in June 7 but not in june 5 and 6
        new_values_june7 = data3 %>%
                anti_join(data2) %>%
                anti_join(data1)
        
        
        #with case number not in june 6 and 7
        cases_in_june5_notin_june67 = data1 %>%
                anti_join(data2, by = c('casenumber')) %>%
                anti_join(data3, by = c('casenumber'))
        
        #with case number in june6 not in june 7
        cases_injune6_notin_june7 = data2 %>%
                anti_join(data3, by = c('casenumber'))
        
        # bind all dataframe
        june7_initial = bind_rows(common_to_all, new_values_june7, cases_in_june5_notin_june67, cases_injune6_notin_june7)
        
        # get the remaining entries from june7
        june7_combined = data3 %>%
                anti_join(june7_initial) %>%
                rbind(june7_initial)
        
        
        # For those without case number that might not be in the june7 combined
        
        #30 patients
        assumed_june5_patients = data1 %>%
                filter(is.na(casenumber)) %>%
                anti_join(june7_combined, by = c('age','sex','nationality','confirminglab','result_date'))
        
        #18 patients
        assumed_june6_patients = data2 %>%
                filter(is.na(casenumber)) %>%
                anti_join(june7_combined, by = c('age','sex','nationality','confirminglab','result_date'))
        
       all_entries = bind_rows(june7_combined, assumed_june5_patients, assumed_june6_patients)
        
       all_entries = all_entries %>%
                distinct() 
                        
       all_entries
}

combined = combine_data(june5, june6, june7)
```

```{r}
clean_combined = function(combined){
        all_entries = all_entries %>%
        mutate(nationality = str_to_title(nationality)) %>%
        mutate(nationality = str_replace(nationality, '^Australia$','Australian'),
               nationality = str_replace(nationality, '^India$','Indian'),
               nationality = str_replace(nationality, 'United Kingdom','British, Uk'),
               nationality = str_replace(nationality, '^Filipino$','Philippine, Filipino'))

        #if case number starts with PH
        all_entries = all_entries %>%
                mutate(nationality = ifelse(str_detect(casenumber, 'PH'),'Philippine, Filipino', nationality )) 
        
        all_entries = all_entries %>%
        mutate(specimen_date = ifelse(year(mdy(specimen_date)) < '2020', result_date, specimen_date))
        
        all_entries = all_entries %>%
        mutate(result_date = ifelse(mdy(result_date)< '2000-01-01',NA, result_date)) %>%
        mutate(result_date = str_replace(result_date, '/[:digit:]{4}', '/2020')) %>%
        mutate(result_date = if_else(mdy(result_date) < '2020-08-01', mdy(result_date), dmy(result_date)))
       
        #return to original format
        all_entries = all_entries %>%
                mutate(result_date = format(result_date, format = '%m/%d/%Y'))
        
        #fix typos
        all_entries = all_entries %>%
        mutate_at(c('comorbidity','fever','cough','cold','sorethroat','difficultyofbreathing','diarrhea','none','quarantined'), str_to_title) 
        
        all_entries
}
combined = clean_combined(combined)
```

```{r}
combined %>%
        write_csv('combined.csv')
```



